Supermartingales and Termination Analysis of Statistical PCF

### 2020-12-28 LICS21/supermartingales-proofs.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper Moved proof of Lemma \ref{lem:rank-Past} to appendix.

>>>>

\rankPast*
\begin{proof}%\rankPast
We first prove (by induction on $n$): for all $n \geq 0$
\[
%\forall n \geq 0 \, . \, 
\expect{Y_n} \leq \expect{Y_0} - \epsilon \cdot \big(\textstyle\sum_{i=0}^{n-1} \mathbb{P}[T > i]\big).
\]
It follows that $\sum_{i=0}^{\infty} \mathbb{P}[T > i]$ converges; hence we have $\lim_{i \to \infty} \mathbb{P}[T > i] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
It then remains to observe: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=0} \mathbb{P}[T > i]$.
\end{proof}

### 2020-12-28 LICS21/supermartingales.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper Moved proof of Lemma \ref{lem:rank-Past} to appendix.

\changed[lo]{Fix a probability space $(\Omega, \calF, \mathbb{P})$, and a filtration $(\calF_n)_{n \geq 0}$. 
(In subsequent applications to termination analysis, $\Omega = \entrosp$, and $\mathbb{P} = \mu_{\entrosp}$.)
Let $T$ be a random variable that takes values in $\mathbb{N} \cup \set{\infty}$.
We call $T$ a \emph{stopping time adapted to $(\calF_n)_{n \geq 0}$} just if $\set{T = n} \in \calF_n$, for all $n \geq 0$.}

\begin{definition}\rm
\begin{enumerate}
\item A sequence of random variables $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~\changed[lo]{(i.e.~for all $A \in \calF_n$, $\int_A \mathbb{P}(\dif \omega) \, Y_{n+1}(\omega) \leq \int_A \mathbb{P}(\dif \omega) \,Y_n(\omega)$).}
\item Let $\epsilon > 0$. 
Given a stopping time $T$ and a supermartingale $(Y_n)_{n \geq 0}$, both adapted to filtration $(\calF_n)_{n \geq 0}$, 
we say that $(Y_n)_{n \geq 0}$ is a \emph{$\epsilon$-ranking supermartingale w.r.t.~$T$} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
{(For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(\omega) := 1$ if $\omega \in A$, and $0$ otherwise.)} \lo{\LaTeX\ does not allow the preceding bracketed sentence to be typeset as a footnote. Strange!}
A \emph{ranking supermartingale} w.r.t.~$T$ is a $\epsilon$-ranking supermartingale w.r.t.~$T$ for some $\epsilon > 0$.
\end{enumerate}
\end{definition}

@@

\begin{lemma}%[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
% \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}
\label{lem:rank-PAST}
Let $(Y_n)_{n \geq 0}$ be a $\epsilon$-ranking supermartingale w.r.t.~the stopping time $T$, then $T < \infty$ a.s., and $\expect{T} \leq \frac{\expect{Y_0}}{\epsilon}$.
\end{lemma}

\begin{proof}
We first prove (by induction on $n$): for all $n \geq 0$
\[
%\forall n \geq 0 \, . \, 
\expect{Y_n} \leq \expect{Y_0} - \epsilon \cdot \big(\textstyle\sum_{i=0}^{n-1} \mathbb{P}[T > i]\big).
\]
It follows that $\sum_{i=0}^{\infty} \mathbb{P}[T > i]$ converges; hence we have $\lim_{i \to \infty} \mathbb{P}[T > i] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
It then remains to observe: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=0} \mathbb{P}[T > i]$.
\end{proof}

>>>>

\changed[lo]{Fix a probability space $(\Omega, \calF, \mathbb{P})$ and a filtration $(\calF_n)_{n \geq 0}$. 
(In subsequent applications to termination analysis, $\Omega = \entrosp$, and $\mathbb{P} = \mu_{\entrosp}$.)
Let $T$ be a random variable that takes values in $\mathbb{N} \cup \set{\infty}$.
We call $T$ a \emph{stopping time adapted to $(\calF_n)_{n \geq 0}$} just if $\set{T = n} \in \calF_n$, for all $n \geq 0$.}

\begin{definition}\rm
\begin{enumerate}
\item A sequence of random variables $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~\changed[lo]{(i.e.~for all $A \in \calF_n$, $\int_A \mathbb{P}(\dif \omega) \, Y_{n+1}(\omega) \leq \int_A \mathbb{P}(\dif \omega) \,Y_n(\omega)$).}
\item Let $\epsilon > 0$. 
Given a stopping time $T$ and a supermartingale $(Y_n)_{n \geq 0}$, both adapted to filtration $(\calF_n)_{n \geq 0}$, 
we say that $(Y_n)_{n \geq 0}$ is a \emph{$\epsilon$-ranking supermartingale w.r.t.~$T$} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
\footnote{(For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(\omega) := 1$ if $\omega \in A$, and $0$ otherwise.)} %\lo{\LaTeX\ does not allow the preceding bracketed sentence to be typeset as a footnote. Strange!}\lo{Now it does. Stranger.}
A \emph{ranking supermartingale} w.r.t.~$T$ is a $\epsilon$-ranking supermartingale w.r.t.~$T$ for some $\epsilon > 0$.
\end{enumerate}
\end{definition}

@@

\begin{restatable}{lemma}{rankPast}
%\begin{lemma}%[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
% \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}
\label{lem:rank-PAST}
Let $(Y_n)_{n \geq 0}$ be a $\epsilon$-ranking supermartingale w.r.t.~the stopping time $T$, then $T < \infty$ a.s., and $\expect{T} \leq \frac{\expect{Y_0}}{\epsilon}$.
%\end{lemma}
\end{restatable}
